---
title: "The Effect of Vitamin C on Tooth Growth in Guinea Pigs"
author: "Derwin McGeary"
date: "09/20/2015"
output: pdf_document
---
### Introduction

This is a brief analysis of the `ToothGrowth` dataset which is distributed with the R programming language.

From the [documentation](http://www.inside-r.org/r-doc/datasets/ToothGrowth) for that dataset:

> The response is the length of odontoblasts (teeth) in each of 10 guinea pigs at each of three dose levels of Vitamin C (0.5, 1, and 2 mg) with each of two delivery methods (orange juice or ascorbic acid).

We wish to know how the delivery method and dosage affect tooth length.

### Exploratory analysis

First we load the data into memory, check the structure, and do a preliminary plot to see what it looks like.
```{r}
data(ToothGrowth)
library(ggplot2)
library(knitr)
str(ToothGrowth)
kable(table(ToothGrowth$dose,ToothGrowth$supp))
summary(ToothGrowth$len)
```

As expected, we have 60 observations (3 dosages, 0.5, 1, and 2; multiplied by 2 supplements, OJ and VC; and 10 guinea pigs per group). We know from the documentation that dosage is in `mg/day`, that `OJ` stands for Orange Juice and `VC` stands for Vitamin C, or ascorbic acid. The units of length are not specified there, but from [Crampton 1947](http://jn.nutrition.org/content/33/5/491.full.pdf) they appear to be microns. For the purposes of this analysis, the units aren't critical.

```{r, echo=FALSE}
qplot(x = ToothGrowth$dose, y=ToothGrowth$len,geom = "smooth", 
      method="lm", 
      colour=ToothGrowth$supp, 
      main="Tooth Growth by Supplement and Dosage",
      xlab="Dosage (mg/day)",
      ylab="Length (microns)") + 
  scale_colour_discrete(name="Supplement") +
  scale_shape_discrete(name="Supplement") +
  geom_point(aes(shape=ToothGrowth$supp))
```

Looking at this graph it appears that there is a positive relationship between dosage and odontoblast length, and that orange juice results in more growth than plain vitamin C.

### Confidence Intervals and Hypothesis Testing

## Dosage and Odontoblast Length
Here we will make three groups by dosage and take two comparisons: 0.5 mg/day versus 1 mg/day; and 1 mg/day versus 2 mg/day.
```{r}
d1 <- subset(ToothGrowth, dose==0.5); d2 <- subset(ToothGrowth, dose==1.0); d3 <- subset(ToothGrowth, dose==2.0)
halfvsone <- t.test(d1$len,d2$len, paired=FALSE,var.equal=FALSE)
onevstwo <- t.test(d2$len,d3$len, paired=FALSE,var.equal=FALSE)
halfvsone; onevstwo
```
## Supplement and Odontoblast Length
Here we will combine all dosages together and group by which supplement was given.
```{r}
s1 <- subset(ToothGrowth, supp=="OJ"); s2 <- subset(ToothGrowth, supp=="VC")
t.test(s1$len,s2$len,paired=FALSE, var.equal=FALSE)
```
### Conclusion
There is a definite positive correlation between increased dosage and larger odontoblasts. This is making the assumption that the distributions are roughly gaussian and independent (`iid`).

There appears to be a relationship between supplement type and growth, but it is not statistically significant at the $\mathrm p=0.5$ level. From our preliminary graph, it may even be important only at lower doses, but that could also be cherry-picking our subsets, a temptation which I will resist.
